Jane, John … Leslie? A Historical Method for Algorithmic Gender Detection

### 2014-09-09 article.Rmd
Tweaks and questions
Minor editing, a few citations, and posed a couple of questions.

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with a range of historical dataset suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also an estimate of how likely that prediction is. For example 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48% were boys. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of male or female.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is unavoidable since the large datasets which make this kind of analysis possible for the past were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define.^[CITE e.g. Margot Canady, The Straight State; works on e.g. Virginia’s definition of race. Shane Landrum on birth certificates. James C. Scott on Seeing Like a State or is that obvious to everyone by now?] Nevertheless, this method is suitable for two reasons. First, researchers are likely to use this method in the aggregate rather than for individuals. The assumption that a name implies male or female (and that those are possible categories to be inferred) is more valid when applied to large groups of people than it does when applied to individuals. Second, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption we have made. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

The remainder of this article is divided into two sections. First, we describe our method in more detail, compare it to existing methods, and explain how to use the method. Second, we apply the method to a case study of the historical profession in order to demonstrate the usefulness of the method.

@@

Our historical method for inferring gender from names may be compared to a method available in the [Natural Language Toolkit](http://www.nltk.org/) for Python.^[Steven Bird, Edward Loper, and Ewan Klein, *Natural Language Processing with Python* (O’Reilly, 2009), ch. 2. Chapter 6 of the NLTK book also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this ahistorical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language. Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpuse provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

The most significant problem with the Kantrowitz names corpus and thus the NLTK implementation is that it assumes that names are timeless. As pointed out above, this makes the corpus unusable for historical purposes. Furthermore the corpus includes some oddities which make it less than suitable. Some names like Abby are overwhelmingly female and some like Bill are overwhelmingly male, but the corpus includes them as both male and female. The Kantrowitz corpus contains only `r kantrowitz_unique`, a mere `r percent(kantrowitz_unique / ssa_unique)` of the `r ssa_unique` names in the Social Security Administration dataset and `r percent(kantrowitz_unique / ipums_unique)` of the `r ipums_unique` names in the census records provided by IPUMS USA. There are therefore many names that it cannot identify. Assuming for the moment that our method provides more accurate results, we estimate that `r percent(kantrowitz_error_rate)` percent of the names in the Kantrowitz corpuse are misclassified or classified as ambigious when a valid prediction could be made.

We mention the Kantrowitz name corpus as implemented in NLTK because the Natural Language Toolkit is rightly regarded as influential for scholarship. Its flaws are also typical of the problems with most other implementations of gender prediction algorithms. The Genderize.io API is, for example, a more sophisticated implementation of gender prediction than the NLTK algorithm. Besides predicting male or female for gender, it also reports the proportion of male or female names, along with a count of the number of uses of the name that it is basing its prediction on. Genderize will also permit the user to customization a prediction for different countries: an extremely important feature. Genderize.io reports that its "database contains 142848 distinct names across 77 countries and 85 languages." Genderize is unsuitable for historical work, however, because it is based only on contemporary data. According to the documentation for its API, "It utilizes big datasets of information, from user profiles across major social networks."^[Quotations are taken from the [Genderize.io website](http://genderize.io), with correspondence from the developer behind the site for confirmation.] It would be anachronistic to apply these datasets to the past, and Genderize.io provides no functionality to filter results chronologically as it does geographically. In addition, exactly what the dataset comprises and how it was gathered is unclear, which keeps scholars from interrogating the value of the source.

R and Python are the most commonly used languages for data analysis among digital humanists. Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. Though options abound for geocoding, for example, there is no satisfactory existing method of gender prediction for historical and humanistic research.

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which contains the predictive algorithm, and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package which contains various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/). The Gender package is affiliated with [rOpenSci](http://ropensci.org/), an initiative that supports reproducible research and open data for scientists using R. The rOpenSci advisory board has provided code review and publicity for the project.

The possibilities for encoding gender from names depends on two things. First, it depends on having a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by states, though the compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year, or range of years since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to make correct predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Ben Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as providing a significant correction to predictions based on Social Security Administration data.

 The Gender package currently includes two datasets which make it suitable for use for United States history since the first federal census in 1790.^[The authors of the package are both American historians, which explains why the initial release of the package focuses on that country. We have plans, however, to extend the underlying data to six other European and North American countries included in the [North Atlantic Population Project](https://www.nappdata.org/napp/) for the nineteenth century. We also welcome contributions of datasets for other regions and times from other scholars.] The first dataset contains names of applicants for Social Security and available from Data.gov. The second dataset contains names gathered in the decennial censuses and available from the IPUMS-USA (Integrated Public Use Microdata Series) service from the Minnesota Population Center at the University of Minnesota.

The Social Security Administration "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[Social Security Administration, "[Baby Names from Social Security Card Applications-National Level Data](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)," Data.gov dataset ID US-GOV-SSA-338 (accessed August 31, 2014). The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910. Social Security Administration, "[Baby Names from Social Security Card Applications-Data by State and District of Columbia](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-)," Data.gov dataset ID US-GOV-SSA-339 (accessed August 31, 2014).] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA has included any name which was used more than five times in a given year; otherwise the dataset is a 100 percent sample. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[CITE history of SSA. On the history of birth registration see Shane Landrum, "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950" (PhD dissertation, Brandeis University, 2014). It is well-known among American historians that Social Security was written to exclude farmworkers with the aim of excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged “a sea change in the [historical] profession with respect to gender” before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck’s report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession. What follows is a quantitative supplement to Lunbeck’s analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. http://www.historians.org/Documents/About%20AHA%20and%20Membership/CWH-Report_5.20.05.pdf

>>>>

Our solution to the "Leslie" problem is to create a software package that combines a predictive algorithm with a range of historical dataset suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also an estimate of how likely that prediction is. For example 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48% were boys. This allows researchers to conduct more nuanced analysis and to determine for themselves what proportion is acceptable for making a prediction of gender.

We recognize that gender is a fluid concept, both today and in the past, and that this method glosses over the complexities of historical definitions of gender and of the relationship between biological sex and gender. In some sense this is unavoidable since the large historical datasets which make this kind of analysis possible were created by state agencies. For the category of gender (as for race) states have offered a limited number of boxes to check, and those boxes extert a tremendous power to define state subjects.^[CITE One of the best-known theorists on gender is Judith Butler, *Gender Trouble* and Butler, *Bodies that Matter*; For the problems with the word "identity" in the humanities and social sciences, see Rogers Brubaker and Frederick Cooper, "Beyond 'Identity'"; Several historians have written about the American state and its drive to impose identities onto subjects: Margot Canady, The Straight State;  Peggy Pascoe, *What Comes Naturally*; Shane Landrum on birth certificates.] Nevertheless, this method is suitable for two reasons. First, researchers are likely to use this method in the aggregate rather than for individuals. The assumption that a name implies male or female (and that those are possible categories to be inferred) is more valid when applied to large groups of people than it does when applied to individuals. **I think this first line of reasoning is dangerous. This leaves us open to critiques of "big data" in general from humanists and theorists: that the process of aggregation is actually really dangerous and that assuming that individual identity and experience can be subsumed under the generalities of the crowd, etc. etc. I think the second reason is strong enough on its own.** Second, this method is neutral as to what the categories male or female actually mean in any given historical context. Researchers are free---indeed, obligated---to make that determination for themselves, and this package provides access to the underlying dataset so that researchers can interrogate every assumption that comes with them. This method does not argue against the complexities of historical understandings of gender, but as with any project it does require researchers who use the method to think through how the data was gathered, what it implies, and how it reflects social and cultural practices.

The remainder of this article is divided into two sections. First, we describe our method in more detail, compare it to existing methods, and explain how to use it. Second, we apply the method to a case study of gatekeeping in the historical profession in order to demonstrate its utility.

@@

Why do scholars need a historical method for inferring gender? Let's start with a comparison. One existing method for predicting gender is available in the [Natural Language Toolkit](http://www.nltk.org/) for Python.^[Steven Bird, Edward Loper, and Ewan Klein, *Natural Language Processing with Python* (O’Reilly, 2009), ch. 2. Chapter 6 of the NLTK book also describes a method of predicting gender from the last letter of the name, but this example is given for the sake of explaining supervised classification and not as an end in itself, so we do not have reason to address this a historical method.] The NLTK is an influential software package for scholarship because it provides an extraordinary range of tools for analyzing natural language. **I think we could develop this even more, if we have information about it. My impression is that it's the "industry standard" for a lot of folks, but it would be nice to give some evidence to then show why our intervention matters, and that the Kantrowitz method isn't just a straw man** Included in the NLTK are two lists of male and female names created by the computer scientist Mark Kantrowitz with assistance from Bill Ross.^[The [original lists of names](http://www.cs.cmu.edu/afs/cs/project/ai-repository/ai/areas/nlp/corpora/names/0.html) are available at the website of the computer science department at Carnegie Mellon University, and the NLTK implementation is described in its [documentation](http://www.nltk.org/howto/corpus.html).] These lists include `r kantrowitz_unique` unique names. Using the Kantrowitz names corpus in the NLTK, one could look up a name like Jane or John and find out that it is male or female. The Kantrowitz corpus provides the list of names.^[The examples below is drawn from our Gender package which implements the Kantrowitz dataset for comparsion's sake, though its use is not recommended for actual research. The Gender package combines the Kantrowitz corpus's two lists into a single list, identifying names that are ambiguous because they appear on both lists as `"either"`.]

@@

The most significant problem with the Kantrowitz names corpus and thus the NLTK implementation is that it assumes that names are timeless. As pointed out above, this makes the corpus problematic for historical purposes. Furthermore the Kantrowitz corpus includes other oddities which make it less suitable. Some names like Abby are overwhelmingly female and some like Bill are overwhelmingly male, but the corpus includes them as both male and female. The Kantrowitz corpus contains only `r kantrowitz_unique` names, a mere `r percent(kantrowitz_unique / ssa_unique)` of the `r ssa_unique` names in the Social Security Administration dataset and `r percent(kantrowitz_unique / ipums_unique)` of the `r ipums_unique` names in the census records provided by IPUMS USA. There are therefore many names that it cannot identify. Assuming for the moment that our method provides more accurate results, we estimate that `r percent(kantrowitz_error_rate)` percent of the names in the Kantrowitz corpus are misclassified or classified as ambigious when a valid prediction could be made. **This isn't immediately obvious to me how this error rate was derived. Do we know why or how Kantrowitz is classifying things as ambiguous? Maybe he's just using a much higher, and theoretically more rigorous, cut-off (say 90%) than our 50% cut-off.**

We mention the Kantrowitz name corpus as implemented in NLTK because the Natural Language Toolkit is rightly regarded as influential for scholarship. Its flaws are also typical of the problems with most other implementations of gender prediction algorithms. The Genderize.io API is, for example, a more sophisticated implementation of gender prediction than the NLTK algorithm. Besides predicting male or female for gender, it also reports the proportion of male or female names, along with a count of the number of uses of the name that it is basing its prediction on. Genderize will also permit the user to customize a prediction for different countries, which is an extremely important feature. Genderize.io reports that its "database contains 142848 distinct names across 77 countries and 85 languages." Genderize is unsuitable for historical work, however, because it is based only on contemporary data. According to the documentation for its API, "It utilizes big datasets of information, from user profiles across major social networks."^[Quotations are taken from the [Genderize.io website](http://genderize.io), with correspondence from the developer behind the site for confirmation.] It would be anachronistic to apply these datasets to the past, and Genderize.io provides no functionality to filter results chronologically as it does geographically. In addition, Genderize.io does not make clear exactly what comprises the dataset and how it was gathered, which keeps scholars from interrogating the value of the source.

R and Python are the most commonly used languages for data analysis among digital humanists.^[**I completely agree, but do we have any evidence for this? Surveys or anything?**] Python's existing packages for gender prediction all implement a method similar to the NLTK or Genderize.io.^[Python library names] To our knowledge, [CRAN](http://cran.rstudio.com/) (the central repository for R packages) does not include any packages dedicated to gender prediction. In two of the most popular data analysis languages in the digital humanities there is currently no satisfactory existing method of gender prediction for historical and humanistic research.

To that end we have created the [Gender](https://github.com/ropensci/gender) package for R which includes both the predictive algorithm and an associated [genderdata](https://github.com/lmullen/gender-data-pkg) package containing various historical datasets. This R implementation is based on an [earlier Python implementation](https://github.com/cblevins/Gender-ID-By-Time) by [Cameron Blevins](http://www.cameronblevins.org/) and [Bridget Baird](http://www.conncoll.edu/directories/emeritus-faculty/bridget-baird/). The Gender package is affiliated with [rOpenSci](http://ropensci.org/), an initiative that supports reproducible research and open data for scientists using R. The rOpenSci advisory board has provided code review and publicity for the project.

The possibilities for inferring gender from names depends on two things. First, it needs a suitable (and suitably large) dataset for the time period and region under study. Unsurprisingly such datasets are almost always gathered in the first instance by government entities, though the compilation and digitization may be undertaken by scholarly researchers. Second, it depends on a suitable algorithm for estimating the proportion of male and female names for a given year or range of years. since often a person cannot be associated with an exact date. It is especially important that the algorithm take into account any biases in the data to formulate more accurate predictions. Development of the R package has had two primary aims. The first is to abstract the predictive algorithm to the simplest possible form so that it is usable for a wide range of historical problems rather than depending on the format of any particular data set. The second has been to provide as many datasets as possible in order to localize the predictions to particular times and places. To that end, [Ben Schmidt](http://benschmidt.org/) contributed a dataset of US Census Data and accompanying code, as well as providing a significant correction to predictions based on Social Security Administration data.

The Gender package currently includes two datasets which make it suitable for studying the United States from the first federal census in 1790 onwards.^[The authors of the package are both American historians, which explains why the initial release of the package focuses on that country. We have plans, however, to extend the underlying data to six other European and North American countries included in the [North Atlantic Population Project](https://www.nappdata.org/napp/) for the nineteenth century. We also welcome contributions of datasets for other regions and times from other scholars.] The first dataset contains names of applicants for Social Security and available from Data.gov. The second dataset contains names gathered in the decennial censuses and available from the IPUMS-USA (Integrated Public Use Microdata Series) service from the Minnesota Population Center at the University of Minnesota.^[**citation?**]

The Social Security Administration "[Baby Names](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)" dataset was created as a result of the Social Security Act of 1935 during the New Deal. ^[Social Security Administration, "[Baby Names from Social Security Card Applications-National Level Data](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-national-level-data)," Data.gov dataset ID US-GOV-SSA-338 (accessed August 31, 2014). The Social Security Administration also provides a related dataset which contains the names of applicants broken down by state and year since 1910. Social Security Administration, "[Baby Names from Social Security Card Applications-Data by State and District of Columbia](http://catalog.data.gov/dataset/baby-names-from-social-security-card-applications-data-by-state-and-district-of-)," Data.gov dataset ID US-GOV-SSA-339 (accessed August 31, 2014).] This dataset contains a list of first names with the number of male and female uses of the name in every year beginning with 1880 and ending with 2012. The SSA list includes any name which was used more than five times in a given year, thereby capturing all but the most infrequently used names from each year. The description "baby names" provided by the SSA is a serious misnomer. When Social Security first became available during the New Deal, its first beneficiaries were adults past or near retirment age. The dataset goes back to 1880, the birth year for a 55 year-old adult when Social Security was enacted. **Okay, so that means we need to tweak some stuff where we use the word "baby" earlier, yes?** Even after 1935, registration at birth for Social Security was not mandatory until 1986. As we will demonstrate below, the way in which the data was gathered requires an adjustment to our predictions of gender.^[CITE history of SSA. On the history of birth registration see Shane Landrum, "The State's Big Family Bible: Birth Certificates, Personal Identity, and Citizenship in the United States, 1840--1950" (PhD dissertation, Brandeis University, 2014). It is well-known among American historians that Social Security was written to exclude farmworkers as a means of also excluding African Americans. In examining the dataset we have not found any reason to believe that this policy noticably skewed the prediction of gender from names.]

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged “a sea change in the [historical] profession with respect to gender” before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck’s report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession. What follows is a quantitative supplement to Lunbeck’s analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period.^[http://www.historians.org/Documents/About%20AHA%20and%20Membership/CWH-Report_5.20.05.pdf]
